---
title: "ASSIGNMENT3"
output: html_document
date: "2023-10-31"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
```

```{r}
college <- read.csv2("college.csv", header = TRUE, sep = ",")
summary(college)
```

```{r}
college$Accept.pct = as.numeric(college$Accept.pct)
college$S.F.Ratio = as.numeric(college$S.F.Ratio)
```

# Problem 1

Apply regression analysis techniques to analyze the relationship among
the observed variables and build a model to predict Graduation Rates
(Grad.Rate). Answer the following questions:

a)  Analyze the distribution of Grad.Rate and discuss if the
    distribution is symmetric, or if you need to apply any
    transformation. [

```{r}
hist(college$Grad.Rate, xlab="Distribution of Graduation rates", prob=TRUE, main="Histogram")
```

Observing above histogram we can say that data is normally distributed
so no need of transformation.

b)  Create scatterplots for Grad.Rate vs each of the independent
    variables. What conclusions can you draw about the relationships
    between Grad.Rate and the independent variables? (No need to include
    the scatterplots in your submission, but you can use correlation
    analysis)

```{r}
correlation = cor(college[ ,sapply(college,is.numeric)])
correlation["Grad.Rate", ]
```

Here we observe that, Elite10,Outstate, Room.Board,Books, PHD, Terminal.
perc.alumni, expend are positively correlated with Grad.Rate while
others are negatively correlated i.e. F.Undergrad , P.Undergrad and
Personal. Outstate and Grad.Rate are postively strongly correlated out
of all.

c)  Build boxplots to evaluate if graduation rates vary by university
    type (private vs public) and by status (elite vs not elite). Discuss
    your findings

```{r}
ggplot(college, aes(x = Private, y = Grad.Rate)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "University Type", y = "Graduation Rate") +
  ggtitle("Graduation Rates by University Type")
```

The garduation rate is more in private as comapared to non-compared. The
universities with yes exhibits its median upper 60's with notable
numbere of outliers. However, universities with no has its median in
lower 60's while and less number of outliers.

```{r}
ggplot(college, aes(x = as.factor(Elite10), y = Grad.Rate)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Graduation Rates by Elite Status",
       x = "Elite Status",
       y = "Graduation Rate")
```

The graduation rate of elite universities is high with around 90% of
median tight IQR. While, non-elite schools have median rate of 70% with
border IQR suggesting more variation. Outliers in this group hint at a
few institutions with substantially lower graduation rates.

d)  Fit a full model (with all independent variables) to predict
    Grad.Rate

```{r}

data = college[ , -which(names(college) == "school")]


```

```{r}

college_fit <- lm(Grad.Rate ~ ., data)
summary(college_fit)
```

About 44.48% variation is shown in graduation rates by the model. The
Adjusted R-squared is 43.46% which is slightly lower. The model is said
to be statistically significant with F-statistics of 43.61 and low
p-value suggest that at least one predictor is significantly affecting
the grad rate.

e)  Does multi-collinearity seem to be a problem here? What is your
    evidence? Compute and analyze the VIF statistics.

```{r}
library(car)
stat<- vif(college_fit)
stat
```

There is no multicollinearity seen from the statistics as all the value
is \<5 which is not a problem. This suggests that the regression
coefficients are reliable, and the model is well-suited for
understanding the factors influencing graduation rates.

f)  Apply TWO variable selection procedures to find an optimal subset of
    independent variables to predict Grad.Rate. You can choose any two
    procedures among the ones we learned in class: backward selection,
    forward selection, adj-R2 , Cp, stepwise, press.

```{r}
step(college_fit, direction = "backward")
```

```{r}
base = lm(Grad.Rate ~ 1, data = data)
step(base , scope = list(upper = college_fit,lower = ~1),direction = "forward")
```

From the above Two variable selection procedures(Backward and Forward),
the minimum step AIC from both the procedure is AIC=3987.16.

g)  Fit a final regression model M1 for Grad.Rate based on the results
    in f). Explain your choice. Write down the expression of the
    estimated model M1

```{r}
M1 = lm(Grad.Rate ~ Outstate + perc.alumni + Accept.pct + 
    P.Undergrad + F.Undergrad + Room.Board + Expend + Personal + 
    Private + PhD + Elite10, data = data)
summary(M1)
```

Choose this model based on AIC value, I have choosed this model, as it
AIC value was minimum. Expression:

Grad.Rate \~ 4.840e+01 + 1.215e-03Outstate + 3.060e-01perc.alumni +
-1.778e+01Accept.pct + -1.963e-03P.Undergrad + F.UndergradF.Undergrad +
1.534e-03Room.Board + -4.465e-04Expend + -1.820e-03Personal +
4.770e+00Private + 8.424e-02PhD + 4.022e+00 \*Elite10

h)  Draw a scatter plot of the studentized residuals against the
    predicted values. Does the plot show any striking pattern indicating
    problems in the regression analysis?

```{r}
plot( fitted(M1), rstudent(M1), main="Predicted vs residuals plot")
abline(a=0, b=0, col='red')
```

The residuals are mostly centered towards zero line and are randomly
scattered. There are few potential outliers nearing or beyond values of
3 and -3 are noticeable but overall we can see a linear realtionship.

i)  Analyze normal probability plot of residuals. Is there any evidence
    that the assumption of normality is not satisfied?

```{r}
qqnorm(rstudent(M1))
qqline(rstudent(M1), col = 2)
```

Here, we observe that majority of residuals are normally distributed,
the deviation tails suggests that certain observations has significantly
larger errors.

j)  Are there any outliers or Influential Points? Compute appropriate
    statistics

```{r}
influencePlot(M1)
```

Yes there are outliers.Points 96 and 378 are outliers, as indicated by
their high Studentized Residuals.

k)  Analyze the R2 value for the final model and discuss how
    welsummary(M1)\$r.squaredl the model explains the variation in
    graduation rates among the universities.

```{r}
summary(M1)$r.squared

```

The R2 value for the final model fitted for Grad.Rate is 0.4431. This
means that the predictors in the model can explain approximately 44.31%
of the variation in graduation rates across universities. The remaining
55.69% of variation is due to factors not accounted for in the model or
to inherent variability.

l)  Draw conclusions on graduation rates based on your regression
    analysis. What are the most important predictors in your model? Does
    your model show a significant difference in graduation rates between
    private and public universities? Do "elite" universities have higher
    graduation rates?

Based on the regression analysis:

Most Important Predictors: - Significant predictors in the model include
"Outstate" , "perc.alumni" , "Accept.pct" , "P.Undergrad" and
"F.Undergrad" , and "Expend" . These factors notable influence on Grad
Rates. Factors with larger coefficients have stronger effects.

Difference Between Private and Public Universities: - Yes, there is a
statistical difference in graduation rates between public and
private.The positive coefficients shows that private have higher
influence than public ones. The p-value is also much smaller than 0.05.

Impact of "Elite" Universities: - The model indicates that "elite"
universities do have higher graduation rates compared to non-elite
universities. The positive coefficient of "Elite10Elite" supports this
conclusion, implying that elite institutions generally achieve better
graduation outcomes.

Conclusion: The analysis says that the significance of outstate, alumni,
accept,pct, institutional type(public or private) and whether it is
elite or not. Graduation rates are impacted by both private or elite
status, showing that the institution has more favorable resources which
supports graduation rate. However, the moderate R2 value indicates there
are other unaccounted factors which affects the graduation date which
model is not able to capture.

# Problem 2

a)  You are asked to build a new regression model that includes the
    following independent variables: Elite10, Accept .pct, Outstate,
    perc.alumni and Expend, together with the interaction effects of
    elite10 with each independent variable. Fit the model and analyze if
    the interaction terms are significant.

```{r}
M2 <- lm(Grad.Rate ~ Elite10 + Accept.pct + Outstate + perc.alumni + Expend +
                 Elite10:Accept.pct + Elite10:Outstate + 
                 Elite10:perc.alumni + Elite10:Expend, data = data)
summary(M2)
```

Here the regression model, including interaction terms, reveals that
about 42.34% of the variation in graduation rates among universities is
explained by factors such as Elite10, Accept.pct, Outstate, perc.alumni,
and Expend, along with their interactions with Elite10. The model is
statistically significant.

b)  Simplify the model and remove interaction terms and additive terms
    that are not significant. Remember that additive terms included in
    interaction terms should not be removed. Write down the expression
    of the final model M2.

```{r}
step(M2, direction = "backward")
```

```{r}
M2_fit = lm(Grad.Rate ~ Elite10 + Accept.pct + Outstate + perc.alumni + 
    Expend + Elite10:Accept.pct + Elite10:Outstate + Elite10:Expend, 
    data = data)
summary(M2_fit)
```

Model(expression):

Grad.Rate = 5.314e+01+ 3.585e+01Elite10 -1.505e+01Accept.pct +
2.322e-03Outstate + 3.334e-01perc.alumni -9.506e-04Expend
-2.164e+01Elite10:Accept.pct -2.253e-03Elite10:Outstate +
1.057e-03Elite10:Expend+e

c)  Analyze the parameter estimates of the fitted model and discuss how
    being an "Elite10" University affects the relationship between
    Graduation Rates and the four predictors Accept.pct, Outstate,
    perc.alumni and Expend.

```{r}
summary(M2_fit)$coefficients[ ,"Estimate"]
```

Graduation rates are affected differently by certain factors when a
university is ranked among the Elite10.

Here we observe that,the negative impact of a higher acceptance rate on
graduation rates is more pronounced in elite universities
Elite10:Accept.pct= -2.163809e+01 . Elite10:Outstate=-2.253419e-03 Elite
universities have less of a positive impact on graduation rates from
higher out-of-state tuition. Elite10:Expend=1.057323e-03 Compared to
non-elite universities, elite universities benefit from higher spending
per student in terms of graduation rates.

\# OPTIONAL: Cross-validation

d)  Apply cross-validation techniques (5-fold cross validation or divide
    dataset into a training and a testing set) to compute how well your
    final model M1 in Part 1 predicts new data. Compute the MAPE (mean
    absolute percentage error) statistic and discuss the results.

```{r}
library(caret)
set.seed(123)
ctrl <- trainControl(method = "cv", number = 5)
cv_M1 <- train(formula(M1), data = data, method = "lm", trControl = ctrl)
summary(cv_M1)
```

```{r}
mape <- function(actual, predicted) {
  mean(abs((actual - predicted) / actual)) * 100
}
```

```{r}
mape_value <- mape(data$Grad.Rate, predict(cv_M1, newdata = data))
mape_value
```

A MAPE of approximately 17.86% indicates that, on average, the model's
predictions of graduation rates are off by 17.86% percentage from the
actual values.

e)  Apply the same cross-validation procedure and compute the MAPE
    statistic for the interaction model M2 computed in Part 2. Compare
    the predictive power of the models M1 and M2 fitted in Part 1 and
    Part 2.

```{r}
cv_M2 <- train(formula(M2_fit), data = data, method = "lm", trControl = ctrl)
summary(cv_M2)
```

```{r}
mape_value2 <- mape(data$Grad.Rate, predict(cv_M2, newdata = data))
mape_value2
```

A MAPE of approximately 18.57% indicates that, on average, the model's
predictions of graduation rates are off by 18.57% percentage from the
actual values.

The model with the lower MAPE is considered to have better predictive
power.Therefore Model 1 has better predictive power.
