---
title: "Assignment_2"
output: html_document
date: "2023-10-16"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
css=read.table("Bankingfull.txt", header=T, sep="\t")
Age=css[,1]
Education=css[,2]
Income=css[,3]
HomeVal=css[,4]
Wealth=css[,5]
Balance=css[,6]


```

```{r}
summary(css)
```

#Problem 1

a)  Create scatterplots to visualize the associations between bank
    balance and the other five variables.

```{r}
plot(Balance,Age, main="Scatterplot of Age vs Account Balance")

```

From above graph. we can observe that the association between Balance
and Age looks fairly linear with few outliers.

```{r}
plot(Balance,Education, main="Scatterplot of Eduction vs Account Balance")

```

From above graph. we can observe that the association between Balance
and Education looks linear with less amount of outliers.

```{r}
plot(Balance,Income, main="Scatterplot of Income vs Account Balance")

```

From above graph. we can observe that the association between Balance
and Income looks linear with few outliers.It seems that both are
strongly correlated.

```{r}
plot(Balance,HomeVal, main="Scatterplot of Account Balance vs HomeVal ")

```

From above graph. we can observe that the association between Balance
and HomeVal looks linear with less outliers.

```{r}
plot(Balance,Wealth, main="Scatterplot of Balance vs Wealth ")
```

From above graph. we can observe that the association between Balance
and Wealth looks linear with no outliers.It seems that wealth and
Balance are strongly correlated.

b)  Compute correlation values of bank balance vs the other variables.
    Interpret the correlation values, and discuss which variables appear
    to be strongly associated.

```{r}
cor(css, method="pearson") 
```

Above data shows all the possible correlations. Here we can observe that
with account balance, wealth and Income are strongly correlated as
(Balance and income) has correlation value of 0.9516845. While, (balance
and Wealth) has correlation value of 0.9487117.We can also say (balance
and Homeval) are strongly correalted with 0.7663871. Other than that,
apart of all the correlations with balance, (wealth and income) with
0.9466654 , (Education and HomeVal) with 0.7535211 are also strongly
correlated

c)  Fit a regression model of balance vs the other five variables (model
    M1). Compute the VIF statistics for each x-variable and analyze
    whether there is a problem of multicollinearity.

```{r}
fullfit <- lm(Balance ~ Age+Education+Income+HomeVal+Wealth, data=css)
summary(fullfit)
```

With above fit, Model 1: Balance=
(-1.071e+04)+(3.187e+02)Age+(6.219e+02)Education+(1.463e-01)Income+(9.183e-03)HomeVal+(7.433e-02)Wealth+e

VIF Statistics

```{r}
library(car)
vif(fullfit)

```

From above tabular data, we observe that there is multicollinearity
problem with wealth and income. Wealth and Income is highly co-linear
with other variables as covariant \>= 10. Wealth has covariant 10.714276
and Income has 14.901724.

d)  Apply your knowledge of regression analysis to define a better model
    M2, and answer the following questions: • Analyze the Coefficient of
    Determination R2 values and the adjusted adj-R2 values for both
    models M1 and M2. Which model has the largest adj-R2 value?

Lets first try removing the multicollinearity covariant:

```{r}
fullfit_2 <- lm(Balance ~ Age+Education+HomeVal, data=css)
summary(fullfit_2)
```

Here, we observe that the R-squared got lowered down to 0.6731 and
adjusted-R2 lowered to 0.6631. Hence, we can say model 1 is better
model. Model 2: Balance=
(-1.933e+04)+(7.194e+02)Age+(3.056e+02)Education+(1.380e-01)HomeVal+e

Now, lets try removing the non significant values x values.

```{r}
fullfit_3 <- lm(Balance ~ Age+Income+Wealth+Education, data=css)
summary(fullfit_3)
```

Here, we observe that R2 and adj-R2 are nearly same but Model 3 has
slightly greater value than model 1. Hence, model 3 is better. Balance=
(-1.243e+04)+(3.251e+02)Age+(6.219e+02)Education+(7.734e+02)Income+(7.299e-02)Wealth+e

-Create residual plots (standardized residuals vs predicted;
standardized residuals vs x-variables; and normal plot of residuals).
Analyze the residual plots to check if the regression model assumptions
are met by the data.

Residuals vs fitted values plot for model 1

```{r}
plot( fitted(fullfit_3), residuals(fullfit_3), main="Predicted vs residuals plot")
abline(a=0, b=0, col='red')

```

From above figure we can say points are randomly scattred around the
zero line.

standardized residuals vs x-variables plot for x-variables

```{r}
plot( Age, residuals(fullfit_3), main="Predicted vs Age plot")
abline(a=0, b=0, col='red')
```

Here we observe that,most of the data is on right side.

```{r}
plot( Income, residuals(fullfit_3), main="Predicted vs Income plot")
abline(a=0, b=0, col='red')
```

From above figure we can say points are randomly scattred around the
zero line.

```{r}
plot( Education, residuals(fullfit_3), main="Predicted vs Education plot")
abline(a=0, b=0, col='red')
```

From above figure we can say points are randomly scattered around the
zero line.

```{r}
plot( HomeVal, residuals(fullfit_3), main="Predicted vs Home values")
abline(a=0, b=0, col='red')
```

From above figure we can say points are randomly scatter around the zero
line.

```{r}
plot( Wealth, residuals(fullfit_3), main="Predicted vs Wealth plot")
abline(a=0, b=0, col='red')
```

```{r}
qqnorm(residuals(fullfit_3))
qqline(residuals(fullfit_3), col = 2)
```

The above graph is normal plot of residuals gives us fairly linear graph
so our normality assumptions are satisfied.

-Analyze if there are any outliers and influential points for your
model. If so, what are your recommendations? Lets find outliers i.e
influential points.

```{r}
influence.measures(fullfit_3)
plot(rstudent(fullfit_3)~hatvalues(fullfit_3)) 
summary(influence.measures(fullfit_3))
```

Here, I observe that there are few outliers and influential points. My
recommendation is that Examine each influential point individually and
Remove an influential point and see if the model results change, or
investigate if influential point is an error or typo.

```{r}
influencePlot(fullfit_3)
```

Observation 38 and 91 stands out as having a relatively high Studentized
Residual (3.635668)and 3.935608 respectively, indicating a potential
outlier. Also, 85 has cooks greater than 1which is 1.39662726 which
shows potential outliers.

-Compute the standardized coefficients and discuss which predictor has
the strongest influence on balance?

```{r}
library(QuantPsyc)
lm.beta(fullfit_3)
```

According to above data, we can say that wealth has highest influence on
balance.

e)  Use the fitted regression model from d) without removal of
    influential points to predict the average bank balance for a
    specific zip code area where there is a plan to open a new branch.
    Census data in that area show the following values: median age is 34
    years, median education is 13 years, median income is \$64,000,
    median home value is \$140,000, median wealth is 160,000. (Note that
    you may not need all these values in your model). Provide predicted
    average bank balance and 95% confidence interval for your estimate.

```{r}
new=data.frame(Age=c(34),Education=c(13),Income=(64000),HomeVal=c(140000),Wealth=c(160000))
predict(fullfit_3, new, interval="prediction", level=0.95) 
               
               
```

Here we can say that predicted average balance is 30575.74 while the 95%
CI interval is (26446.1,34705.38)

# Problem 2

```{r}
css_1=read.table("pgatour2006_small.csv", header=T, sep=",")
Name=css_1[,1]
PrizeMoney=css_1[,2]
DrivingAccuracy=css_1[,3]
GIR=css_1[,4]
PuttingAverage=css_1[,5]
BirdieConversion=css_1[,6]
PuttsPerRound=css_1[,7]

```

a)  Create scatterplots to visualize the associations between PrizeMoney
    and the other five variables. Discuss the patterns displayed by the
    scatterplot. Do the associations appear to be linear?

```{r}
plot(PrizeMoney,DrivingAccuracy, main="Scatterplot of Prize money vs Driving Accuracy")
```

From above plot, we can observe that association between Prize money and
Driving Accuracy the data is not linearly associated.

```{r}
plot(PrizeMoney,GIR, main="Scatterplot of Prize money vs GIR")
```

From above plot, we can observe that association between Prize money and
GIR the data is not linearly associated.

```{r}
plot(PrizeMoney,PuttingAverage, main="Scatterplot of Prize money vs Putting Average")
```

From above plot, we can observe that association between Prize money and
Putting Average the data is not linearly associated.

```{r}
plot(PrizeMoney,BirdieConversion, main="Scatterplot of Prize money vs Birdie Conversion")
```

From above plot, we can observe that association between Prize money and
Birdie conversion the data is not linearly associated.

```{r}
plot(PrizeMoney,PuttsPerRound, main="Scatterplot of Prize money vs Putts Per Round")
```

From above plot, we can observe that association between Prize money and
PuttsPerRound the data is not linearly associated.

So we can say that, the association of Price Money with other five
variables is not linear.

b)  Analyze distribution of PrizeMoney, and discuss if the distribution
    is symmetric or skewed.

```{r}
hist(PrizeMoney, xlab="Prize Money", prob=TRUE, main="Histogram") 
library(psych)
describe(PrizeMoney)
```

Here from the figure we observe that, most of the data is in the left
side of the graph so we can say that distribution is right skewed. Other
than that here, mean is greater than median which also supports that
data distribution is right skewed.

c)  Apply a log transformation to PrizeMoney and compute the new
    variable ln_Prize=log(PrizeMoney). Analyze distribution of ln_Prize,
    and discuss if the distribution is symmetric or skewed.

```{r}
hist(log(PrizeMoney), xlab="Prize Money", prob=TRUE, main="Histogram") 
library(psych)
describe(log(PrizeMoney))
```

After applying log transformation, we can see that the data distribution
is approximately symmetric and normal.While here, mean and median are
nearly same which supports that the distribution is symmetric.

d)  Fit a regression model of ln_Prize using the remaining predictors in
    your dataset. Apply your knowledge of regression analysis to define
    a valid model to predict ln_Prize. Hint: use scatter plots and
    correlation

```{r}
coln<-css_1[,2:7]
cor(coln,method='pearson')
```

```{r}
fullpga <- lm(log(PrizeMoney) ~ DrivingAccuracy+GIR+PuttingAverage+BirdieConversion+PuttsPerRound, data=css_1)
summary(fullpga)
```

Above is a model design including all variables. Model_1:
log(PrizeMoney)=0.466245+(0.004063)DrivingAccuracy+(0.257773)GIR+(11.745503)PuttingAverage+(0.175061)BirdieConversion-(1.133206)PuttsPerRound+e
Here,we got R-squared: 0.5264, Adjusted R-squared: 0.5108

```{r}
vif(fullpga)
```

Here we can see all are \<10 so there is no multicollinearity problem.

-If necessary remove not significant variables. Remember to remove one
variable at a time (variable with largest p-value is removed first) and
refit the model, until all variables are significant.

Lets remove non significant variables

```{r}
newfullpga <- lm(log(PrizeMoney) ~ GIR+BirdieConversion+PuttsPerRound, data=css_1)
summary(newfullpga)
```

Here we removed the non-significant variables and refit the model and
the model we got is: log(PrizeMoney)=11.78325+0.23571 (GIR)+0.11702
(BirdieConversion)-0.68946(PuttsPerRound )+e

Here we got R-squared: 0.5161, Adjusted R-squared: 0.5034. We can see
first model was better than this.

Analyze residual plots to check if the regression model is valid for
your data.

```{r}
plot( fitted(fullpga), residuals(fullpga), main="Predicted vs residuals plot")
abline(a=0, b=0, col='red')
```

The graph describes that, data is randomly scattered.

```{r}
plot( DrivingAccuracy, residuals(fullpga), main="Predicted vs Driving Accuracy plot")
abline(a=0, b=0, col='red')

```

```{r}
plot( GIR, residuals(fullpga), main="Predicted vs GIR plot")
abline(a=0, b=0, col='red')
```

```{r}
plot( PuttingAverage, residuals(fullpga), main="Predicted vs Putting Average plot")
abline(a=0, b=0, col='red')
```

```{r}
plot( BirdieConversion, residuals(fullpga), main="Predicted vs Birdie Conversion plot")
abline(a=0, b=0, col='red')
```

```{r}
plot( PuttsPerRound, residuals(fullpga), main="Predicted vs PuttsPerRound")
abline(a=0, b=0, col='red')
```

From the residual vs fitted and the residuals vs all the x-values we
observe that the distribution is random and scattered.

```{r}
qqnorm(residuals(fullpga))
qqline(residuals(fullpga), col = 2)

```

Hence, random distribution gives us fairly linear points in plot . Hence
assumption of normality satisfied.

-Analyze if there are any outliers and influential points. If there are
points in the dataset that need to be investigated, give one or more
reason to support each point chosen.

```{r}
influence.measures(fullpga)
plot(rstudent(fullpga)~hatvalues(fullpga)) 
summary(influence.measures(fullpga))
```

Observing above graph we can say there is only one outlier.Influential
points can have a significant impact on the regression model's
parameters and predictions. These points may be outliers in the dataset,
having extreme values for some predictors.They may represent cases that
deviate from the overall pattern observed in the data. I choose one
point as an outlier as it was High Deleted Studentized Residuals
(outside (-3,3) band).Other than that if the point has High leverage hat
hii value (\> 0.5) or High Cook's D distance (\>1) than that is
considered as outlier or influential point.

```{r}
influencePlot(fullpga)
```

e)  Interpret the regression coefficients in the final model to answer
    the following question: How does an increase in 1% for GIR affect
    the average Prize money?

```{r}
coef(fullpga)["GIR"]
exp(coef(fullpga)["GIR"])

```

Here, coefficient estimate for GIR is 0.2577731. This represents the
change in Prize Money when we make change of one unit increase in
predictor variable, when all the other variables are constant. So, 1%
change in GIR will make 1.294045 change to Prize Money

f)  Compute the prediction and 95% prediction interval for average prize
    money for a player that has a GIR of 67%, driving accuracy of 64%,
    putting average of 1.77, Birdie Conversion of 28% and 29.16 average
    putts per round.

```{r}
new=data.frame(DrivingAccuracy=c(64),GIR=c(67),BirdieConversion=c(28),PuttingAverage=c(1.77),PuttsPerRound=c(29.16))
predict(fullpga, new, interval="prediction", level=0.95) 
               
```

```{r}
exp(predict(fullpga, new, interval="prediction", level=0.95) )
```

This model prediction for a player that has a GIR of 67%, driving
accuracy of 64%, putting average of 1.77, Birdie Conversion of 28% and
29.16 average putts per round gives 95% prediction interval
(10741.53,163764.2) and average Price money 41941.37.
